## Setup

```{r}
#| label: setup
library(survival)
library(survminer)
library(tidyverse)
library(data.table)
library(broom)
```


## Load data
```{r}
addicts <- fread("https://web1.sph.emory.edu/dkleinb/allDatasets/surv2datasets/addicts.dat", header = F, sep=" ", skip = 19) |> 
  rename_with(~ c("ID", "Clinic", "status", "survt", "prison","dose"))

anderson <- fread("https://web1.sph.emory.edu/dkleinb/allDatasets/surv2datasets/anderson.dat") |>
  mutate(ID = row_number()) |>
  select(ID, everything()) |> 
  rename_with(~c("Subj", "Survt", "Relapse", "Sex", "logWBC", "Rx")) |> 
  mutate(Group = case_when(logWBC <= 2.3 ~ "Low",
                           logWBC < 3 ~ "Medium",
                           .default = "High")) 
anderson|>
  janitor::tabyl(Group)
```

```{r}
# split data
addicts.cp <- addicts |> 
  survSplit(cut = survt[status == 1],
            end = "survt",
            event = "status",
            start = "start",
            id = "id")
            
```
```{r}
view(veteran)
vet2 <- survSplit(Surv(time, status) ~., veteran,
                   cut=c(60, 120), episode ="timegroup")

addicts.cp <- survSplit(addicts,cut=addicts$survt[addicts$status==1],
                        end="survt", event= "tatus",start="start",id= "id")
```

```{r}
testdata <- data.frame(id = 1:3,
                       syear = c(1945, 1967, 1950),
                       eyear = c(1950, 1970, 1965),
                       status = c(1, 1, 2),
                       age = c(45, 50, 40),
                       group = c("A", "B", "A"))
testdata
```

```{r}
#survSplit by year, each year is a new row
survSplit(Surv(syear, eyear, status) ~ ., 
          testdata, 
          cut = 1945:1970,
          event = "status")
```

```{r}
survSplit(
  testdata,
  cut = 1945:1970,
  end = "eyear",
  event = "status",
  start = "syear",
  episode = "episode"
)
```

```{r}
expand_data <- function(data) {
  new_rows <- list()
  
  for (i in 1:nrow(data)) {
    id <- rep(data$id[i], data$eyear[i] - data$syear[i] + 1)
    year <- seq(data$syear[i], data$eyear[i])
    age <- data$age[i] + (year - data$syear[i])
    status <- ifelse(year == data$eyear[i], data$status[i], 0)
    group <- rep(data$group[i], length(id))
    
    new_rows[[i]] <- data.frame(id, year, age, status, group)
  }
  
  expanded_data <- do.call(rbind, new_rows)
  return(expanded_data)
}
```


```{r}
expand_data <-
  function(data,
           id_var,
           start_var,
           end_var,
           increment_var = NULL,
           status_var = NULL,
           by = "year",
           increment = 1,
           status_code = 0) {
    new_rows <- list()
    
    for (i in 1:nrow(data)) {
      id <-
        rep(data[[id_var]][i], (data[[end_var]][i] - data[[start_var]][i]) / increment + 1)
      
      if (!is.null(increment_var)) {
        time_values <-
          seq(data[[start_var]][i], data[[end_var]][i], by = data[[increment_var]][i])
      } else {
        time_values <-
          seq(data[[start_var]][i], data[[end_var]][i], by = increment)
      }
      
      if (increment == -1) {
        time_values <- rev(time_values)
      }
      
      age <-
        data[[start_var]][i] + (time_values - data[[start_var]][i])
      
      if (!is.null(status_var)) {
        status <-
          ifelse(time_values == data[[end_var]][i], data[[status_var]][i], status_code)
      } else {
        status <- rep(status_code, length(id))
      }
      
      other_vars <-
        setdiff(names(data),
                c(id_var, start_var, end_var, increment_var, status_var))
      other_data <-
        lapply(other_vars, function(var)
          rep(data[[var]][i], length(id)))
      
      new_row <-
        cbind(
          id = id,
          time_values = time_values,
          age = age,
          status = status,
          other_data
        )
      new_rows[[i]] <- as.data.frame(new_row)
    }
    
    expanded_data <- do.call(rbind, new_rows)
    return(expanded_data)
  }

# Example usage
testdata |>
  expand_data(
    id_var = "id",
    start_var = "syear",
    end_var = "eyear",
    increment_var = "age",
    status_var = "status",
    increment = 1,
    status_code = 0
  )
```

```{r}
expand_dataset <- function(df, start_time_var, end_time_var, by = "year",
                           add_vars = NULL, add_values = 1,
                           status_vars = NULL, status_default = 0) {
  
  # Expand by time duration
  time_expanded <- df %>%
    tidyr::expand(nesting(!!!syms(setdiff(names(.), c(start_time_var, end_time_var)))),
                  !!sym(start_time_var) := seq(first(!!sym(start_time_var)),
                                               last(!!sym(end_time_var)),
                                               by = by))
  
  # Add values to specified variables
  if (!is.null(add_vars)) {
    time_expanded <- time_expanded %>%
      group_by(across(-!!sym(start_time_var))) %>%
      mutate(across(all_of(add_vars), ~ . + seq(add_values * (n() - 1), 0, by = -add_values)))
  }
  
  # Assign status values
  if (!is.null(status_vars)) {
    time_expanded <- time_expanded %>%
      mutate(across(all_of(status_vars), ~ ifelse(!!sym(start_time_var) == last(!!sym(end_time_var)),
                                                   ., status_default)))
  }
  
  return(time_expanded)
}
```

```{r}
testdata |> 
  expand_dataset(start_time_var = "syear", end_time_var = "eyear", by = 1, 
                 add_vars = "age", add_values = 1, status_vars = "status", status_default = 0) |> 
  view()
```

```{r}
group1 <- c(6, 6,6,7,10,
13, 16,22,23,
6, 9, 10, 11,
17, 19, 20, 25,
32, 32, 34, 35)
group2 <- c(1, 1, 2, 2,3,
4,4,5,5,
8,8,8, 8,
11, 11, 12,12,
15, 17, 22, 23)

9/sum(group1)
21/sum(group2)
```
```{r}
survdat <- read.csv("data/Survival.csv")

survdat <- survdat |> 
  mutate(censored = ifelse(Turnover == 1, 1, 0))

survdat |> 
  ggplot(aes(x = LOS)) +
  geom_histogram(bins = 10, col = "purple", fill = "white")
```
```{r}
survdat |>
  ggplot(aes(x = LOS)) +
  geom_histogram(bins = 10, col = "purple", fill = "white") +
  facet_wrap(~ Turnover)
```
### survival datasets

aml (cancer)             Acute Myelogenous Leukemia survival data
bladder (cancer)         Bladder Cancer Recurrences
bladder1 (cancer)        Bladder Cancer Recurrences
bladder2 (cancer)        Bladder Cancer Recurrences
cancer (lung)            NCCTG Lung Cancer Data
capacitor (reliability)

```{r}
aml
bladder
bladder1
bladder2
cancer
capacitor
?lung

```

```{r}
AIDSsurv <- data.frame(
  Time = c(2, 3, 6, 6, 7, 10, 15, 15, 16, 27, 30, 32),
  Death = c(1, 0, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1)
)
AIDSsurv
```


## Model

### Kaplan-Meier
```{r}
kp<- AIDSsurv |> 
  survfit(Surv(Time, Death) ~ 1, data = _) 
kp
summary(kp)
tidy(kp)
```

```{r}
ggsurvplot(kp, data = AIDSsurv, risk.table = TRUE, pval = TRUE, conf.int = TRUE, ggtheme = theme_bw(),surv.median.line = "hv", legend.labs = c("AIDS"))
```

```{r}
kp.aml <- aml |> 
  survfit(Surv(time, status) ~ x, data = _)
kp.aml
tidy(kp.aml)
```

```{r}
kp.aml |> 
  ggsurvplot(risk.table = TRUE, pval = TRUE, conf.int = F, ggtheme = theme_bw(),
             surv.median.line = "hv", legend.labs = c("Maintained", "Non-aintained"), 
             data = aml)
```

## Expontential model

```{r}

```


```{r}
survdat |> 
  survfit(Surv(LOS, censored) ~ 1, data = _) |> 
  tidy()
```
```{r}
# Load the survival package
# install.packages("survival")
# library(survival)

# Fit Cox PH model using the lung dataset
cox_model <- coxph(Surv(time, status) ~ age + sex + ph.karno, data = lung)

# Generate survival curves
surv_curve <- survfit(cox_model)

# Print summary of survival curves
summary(surv_curve)

# Plot the survival curve
plot(surv_curve, main = "Survival Curve for Lung Cancer Patients", xlab = "Time (Days)", ylab = "Survival Probability")

ggsurvplot(surv_curve, data = lung, risk.table = TRUE, pval = TRUE, conf.int = TRUE, ggtheme = theme_bw())
lung
```

```{r}
# Load the survival package
# install.packages("survival")
# library(survival)

# coxph model with lung dataset
lung |> 
  coxph(Surv(time, status) ~ age + sex + ph.karno, data = _) |>
  survfit() |> 
  ggsurvplot_group_by(data = lung, risk.table = TRUE, pval = TRUE, conf.int = TRUE, group.by = "sex",  ggtheme = theme_bw())
```

```{r}
library(survival)
fit <- survfit(Surv(time, status) ~ sex, data = colon )

# Visualize: grouped by treatment rx
#:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
ggsurv.list1 <- ggsurvplot_group_by(fit, colon, group.by = "rx", risk.table = TRUE,
                                 pval = TRUE, conf.int = TRUE, palette = "jco")
names(ggsurv.list)


# Visualize: grouped by treatment rx and adhere
#:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
ggsurv.list2 <- ggsurvplot_group_by(fit, colon, group.by = c("rx", "adhere"),
                                 risk.table = TRUE,
                                 pval = TRUE, conf.int = TRUE, palette = "jco")

ggsurv.list1; ggsurv.list2
```

## Cox PH model
```{r}
model_anderson <- anderson |> 
  coxph(Surv(Survt, Relapse) ~ logWBC + Rx + Sex, data = _)
model_anderson |>
  summary()
```

```{r}
test2 <- data.frame(start=c(1,2,5,2,1,7,3,4,8,8), 
              stop=c(2,3,6,7,8,9,9,9,14,17), 
              event=c(1,1,1,1,1,1,1,0,0,0), 
              x=c(1,0,0,1,0,1,1,1,0,0)) 
summary(coxph(Surv(start, stop, event) ~ x, test2))
```
```{r}
test1 <- data.frame(time=c(4,3,1,1,2,2,3), 
              status=c(1,1,1,0,1,1,0), 
              x=c(0,2,1,1,1,0,0), 
              sex=c(0,0,0,0,1,1,1))
coxph(Surv(time, status) ~ x + strata(sex), test1) 
coxph(Surv(time, status) ~ x + sex, test1) 
```

```{r}
anderson |> 
  coxph(Surv(Survt, Relapse) ~ Rx + logWBC, data = _, method = "efron") |> 
  summary()
anderson |> 
  coxph(Surv(Survt, Relapse) ~ Rx + logWBC, data = _, method = "breslow") |> 
  summary()
anderson |> 
  coxph(Surv(Survt, Relapse) ~ Rx + logWBC, data = _, method = "exact") |> 
  summary()
  
```
```{r}
anderson |> 
  filter(Sex == 0) |> 
  coxph(Surv(Survt) ~ Rx + logWBC, data = _) |> 
  summary()
  
```
```{r}
anderson |> 
  filter(Sex == 1) |> 
  coxph(Surv(Survt) ~ Rx + logWBC, data = _) |> 
  summary()
  
```

```{r}
anderson |> 
  coxph(Surv(Survt, Relapse) ~ Rx + logWBC + Rx:Sex + logWBC:Sex, data = _) |> 
  summary()
  
```
```{r}
anderson |> 
  coxph(Surv(Survt, Relapse) ~ logWBC + Rx + Sex, data = _, method = "breslow") |> 
  summary()

anderson |> 
  coxph(Surv(Survt, Relapse) ~ logWBC + Rx + Sex, data = _, method = "breslow") |> 
  cox.zph() 
```
## Class 16: Cox PH model
```{r}
anderson |> 
  coxph(Surv(Survt, Relapse) ~ Rx, data = _, method = "breslow") |> 
  summary()

mod_anderson2 <- anderson |> 
  coxph(Surv(Survt, Relapse) ~ Rx + logWBC, data = _, method = "breslow")

anderson |> 
  coxph(Surv(Survt, Relapse) ~ Rx * logWBC, data = _, method = "breslow") |> 
  summary()
```
```{r}
# plot survival curves for coxph model, compare by Sex
mod_anderson2 <- anderson |> 
  coxph(Surv(Survt, Relapse) ~ Rx + logWBC + Sex, data = _, method = "breslow")

plotdata <- anderson |> 
  summarise(logWBC = mean(logWBC),
            .by = c(Rx, Sex))

mod_anderson2 |> 
  survfit(newdata = plotdata) |> 
  ggsurvplot(data = plotdata, 
             surv.median.line = "hv")

newdata <- data.frame(Rx = 0:1,
                      Sex = 0:1,
                      logWBC = mean(anderson$logWBC,))

mod_anderson2 |>
  survfit(newdata = newdata) |> 
  ggsurvplot(data = newdata, 
             censor = F,
             risk.table = TRUE, 
             risk.table.y.text.col = TRUE,
             legend.labs = c("Sex = Male", "Sex = Female"),
             #legend.labs = c("Rx = New Treatment", "Rx = Standard Treatment"),
             ggtheme = theme_bw())
mod_anderson2 |>
  survfit(newdata = newdata) |> 
  ggsurvplot(data = newdata, 
             censor = F,
             risk.table = TRUE, 
             risk.table.y.text.col = TRUE,
             #legend.labs = c("Sex = Male", "Sex = Female"),
             legend.labs = c("Rx = New Treatment", "Rx = Standard Treatment"),
             ggtheme = theme_bw())
```

```{r}
mod_anderson2 |> 
  survfit(newdata = plotdata) |> 
  ggsurvplot(data = anderson, 
             surv.median.line = "hv")
```
```{r}
mod_anderson2 |> 
  survfit(newdata = plotdata) |> 
  ggsurvplot(data = anderson, 
             surv.median.line = "hv",
             legend.labs = c("Rx = 0, Sex = 1",
                             "Rx = 0, Sex = 0",
                             "Rx = 1, Sex = 1",
                             "Rx = 1, Sex = 0"))
```
```{r}
anderson
```
```{r}
fit <- survfit( Surv(time, status) ~ sex, data = colon )
ggsurvplot_group_by(fit, colon, group.by = "rx",
                palette = "jco", pval = TRUE)
```


```{r}
mod1 <- addicts|> 
  coxph(Surv(survt, status) ~ prison + dose + Clinic, data = _, method = "breslow")
mod1
mod2 <- addicts|> 
  coxph(Surv(survt, status) ~ prison * Clinic + dose * Clinic, data = _, method = "breslow")
mod2
```
```{r}
# Create a function to do the likelihood ratio test
lrt.surv <-  function(mod.reduced, mod.full, df) {
  lrts <-  (-2) * (mod.reduced$loglik[2] - mod.full$loglik[2])
  pvalue <-  1 - pchisq(lrts, df)
  return(pvalue)
}

lrt.surv(mod1, mod2, 2)
```


```{r}
model_mtcars <- mtcars |> 
  lm(mpg ~ wt + cyl, data = _)

names(model_mtcars)
```
```{r}
model_mtcars$call
```

## Running a Stratified Cox Model
```{r}
addicts |> 
  coxph(Surv(survt, status) ~ prison + dose + strata(Clinic), data = _, method = "breslow") |> 
  summary()
```
```{r}
addicts |> 
  coxph(Surv(survt, status) ~ prison + dose + prison:Clinic + dose:Clinic + strata(Clinic), data = _, method = "breslow") |> 
  summary()
```

```{r}
exp(0.388788 + 2 * -0.035145)
```
```{r}
anderson |> 
  coxph(Surv(Survt, Relapse) ~ Rx * logWBC, data = _, method = "breslow") |> 
  summary()
```

```{r}
ovarian
```


## Example of Cancer Data
```{r}
cancer
```
```{r}
cancer <- cancer |> 
  drop_na()
model_cancer <- cancer |> 
  coxph(Surv(time, status) ~ age + ph.karno + pat.karno + meal.cal + wt.loss + sex, data = _, method = "breslow")
model_cancer|> 
  summary()
```
```{r}

plot_data_cancer <- data.frame(
  age = mean(cancer$age),
  ph.karno = mean(cancer$ph.karno),
  pat.karno = mean(cancer$pat.karno),
  meal.cal = mean(cancer$meal.cal),
  wt.loss = mean(cancer$wt.loss),
  sex = 1:2
)

model_cancer |> 
  survfit(newdata = plot_data_cancer) |>
  ggsurvplot(data = cancer,
             censor = F,
             risk.table = F, 
             pval = TRUE, 
             legend.labs = c("Sex = Male", "Sex = Female"),
             ggtheme = theme_bw())
```


## Class 17: Review Cox PH Model- PH assumption; Checking PH: 3 methods

### Cox PH Model
```{r}
anderson |> 
  coxph(Surv(Survt, Relapse) ~ Rx * logWBC, data = _, method = "breslow") |> 
  summary()
```
### PH assumption check

```{r}
addicts |> 
  head(3)
kmfit1 <- addicts |> 
  survfit(Surv(survt, status) ~ 1, data = _)

summary(kmfit1)
```

```{r}
kmfit2 <- addicts |> 
  survfit(Surv(survt, status) ~ Clinic+ strata(prison), data = _) 
summary(kmfit2)
```


```{r}
addicts |> 
  survdiff(Surv(survt, status) ~ Clinic+ strata(prison), data = _)
addicts |> 
  survdiff(Surv(survt, status) ~ Clinic+ strata(prison), rho = 1, data = _)
```
```{r}
kmfit2 |> 
  ggsurvplot(fun = "cloglog")
```

```{r}
kmfit3 <- summary(kmfit2)
kmfit3$strata
kmfit4 <- data.frame(kmfit3$strata, kmfit3$time, kmfit3$surv) |> 
  rename_with(~ c("clinic", "time", "survival"))
```

```{r}
clinic1 <-  kmfit4 |> 
  filter(str_detect(clinic, "Clinic=1"))
clinic2 <- kmfit4 |> 
  filter(str_detect(clinic, "Clinic=2"))
```

```{r}
plot(
  clinic1$time,
  log(-log(clinic1$survival)),
  xlab = "survival time in days",
  ylab = "log-log survival",
  xlim = c(0, 800),
  col = "red",
  type = 'l',
  lty = "solid",
  main = "log-log
curves by clinic"
)

par(new = T)

plot(
  clinic2$time,
  log(-log(clinic2$survival)),
  axes = F,
  xlab = "survival time in days",
  ylab = "log-log survival",
  col = "purple",
  type = 'l',
  lty = "dashed"
)

legend(
  "bottomright",
  c("Clinic 1", "Clinic 2"),
  lty = c("solid", "dashed"),
  col = c
  ("red", "purple")
)

par(new = F)
```

## Assessing the PH Assumption with a stastical test
```{r}
mod1 <- addicts |> 
  coxph(Surv(survt, status) ~ prison + dose + Clinic, data = _, method = "breslow")
cox.zph(mod1)
```
```{r}
mod.zph <- cox.zph(mod1, transform = rank)
ggcoxzph(
  mod.zph,
  resid = TRUE,
  se = TRUE,
  df = 4,
  nsmo = 40,
  point.col = "red",
  point.size = 1,
  point.shape = 19,
  point.alpha = 1,
  ggtheme = theme_survminer()
)
```

```{r}
mod.anderson <- anderson |> 
  coxph(Surv(Survt, Relapse) ~ Rx + logWBC + Sex, data = _, method = "breslow")

plotdata1 <- anderson |> 
  summarise(logWBC = mean(logWBC),
            Rx = mean(Rx),
            .by = Sex)

plotdata2 <- data.frame(logWBC = mean(anderson$logWBC),
                        Rx = 0:1)

mod.anderson |>
  survfit(newdata = plotdata1) |> 
  ggsurvplot(data = anderson,
             fun = "cloglog")
```
```{r}
```


## Obtaining cox-adjusted survival curves
```{r}
mod1 <- addicts |> 
  coxph(Surv(survt, status) ~ prison + dose + Clinic, data = _, method = "breslow")

pattern1 <- data.frame(
  prison = 0,
  dose = 70,
  Clinic = 1:2
)
pattern1
summary(survfit(mod1, newdata = pattern1))
```
```{r}
mod1 |> 
  survfit(newdata = pattern1) |> 
  ggsurvplot(data = addicts, 
             conf.int = F,
             legend.labs = c("Clinic = 1", "Clinic = 2"),
             surv.median.line = "hv",
             ggtheme = theme_bw())
```
```{r}
mod3 <- addicts |> 
  coxph(Surv(survt, status) ~ prison + dose + strata(Clinic), data = _, method = "breslow")
```
```{r}
pattern2 <- data.frame(
  prison = .46,
  dose = 60.4,
  Clinic = 1:2
)
pattern2
summary(survfit(mod3, newdata = pattern2))
```

### plot 
```{r}
mod3 |> 
  survfit(newdata = pattern2) |>
  ggsurvplot(data = addicts, 
             conf.int = F,
             legend.labs = c("Clinic = 1", "Clinic = 2"),
             surv.median.line = "hv",
             ggtheme = theme_bw())
```
```{r}
mod3 |> 
  survfit(newdata = pattern2) |> 
  ggsurvplot(fun = "cloglog",
             data = pattern2,
             # xlim = c(10, 1000),
             # ylim = c(-5, 1),
             ggtheme = theme_bw())
```
```{r}
mod4 <- anderson |> 
  coxph(Surv(Survt, Relapse) ~ Rx + logWBC, data = _)
mod5 <- anderson |> 
  coxph(Surv(Survt, Relapse) ~ Sex + logWBC, data = _)
mod6 <- anderson |> 
  coxph(Surv(Survt, Relapse) ~ Sex + logWBC + Rx, data = _)
```
```{r}
plotdata <- data.frame(
  Rx = 0:1,
  logWBC = mean(anderson$logWBC)
)
plotdata1 <- anderson |> 
  summarise(logWBC = mean(logWBC), 
            .by = Rx)
plotdata2 <- anderson |> 
  summarise(logWBC = mean(logWBC), 
            .by = Sex)
plotdata3 <- anderson |> 
  summarise(logWBC = mean(logWBC), 
            .by = c(Sex, Rx))
mod4 |> 
  survfit(newdata = plotdata1) |>
  ggsurvplot(data = anderson,
             conf.int = F,
             linetype = 1:2,
             fun = "cloglog")
mod5 |> 
  survfit(newdata = plotdata2) |>
  ggsurvplot(data = anderson,
             conf.int = F,
             linetype = 1:2,
             fun = "cloglog")
mod6 |> 
  survfit(newdata = plotdata3) |>
  ggsurvplot(data = anderson,
             conf.int = F,
             fun = "cloglog")
mod6 |> 
  survfit(newdata = plotdata2) |>
  ggsurvplot(data = anderson,
             conf.int = F,
             fun = "cloglog")
```
```{r}
data()
jasa1 |> 
  mutate(Over40 = factor(ifelse(age + 48 > 40, 1, 0)),
         Time = stop - start) |>
  coxph(Surv(Time, event) ~ Over40, data = _, method = "breslow") |> 
  summary()
```

```{r}

```
## Class 19: Time Dependent Variables; Extended Cox Model

```{r}
veteran
```
```{r}
veteran |> 
  coxph(Surv(time, status) ~ trt + celltype + karno + diagtime + age + prior, data = _, method = "breslow") |> 
  summary()
```
### cut data

```{r}
#survSplit by year, each year is a new row
survSplit(Surv(syear, eyear, status) ~ ., 
          testdata, 
          cut = 1945:1970,
          event = "status") |>
  view()
```

```{r}
expanded_data <- survSplit(
  testdata,
  cut = 1945:1970,
  end = "eyear",
  event = "status",
  start = "syear",
  episode = "episode"
)
addicts
```



```{r}
survSplit(
  addicts,
  cut = 0:1,
  end = "status",
  event = "status",
  start = "status"
)
```
```{r}
addicts.cp <- addicts |> 
  survSplit(Surv(survt, status) ~ .,
            data = _,
            cut = addicts$survt[addicts$status == 1],
            event = "status",
            start = "start",
            end = "survt") |> 
  mutate(logt_dose = dose * log(survt))
```
```{r}
addicts.cp |> 
  select(ID, start, survt, status, dose, logt_dose) |>
  filter(ID == 114)
```
```{r}
addicts.cp |> 
  coxph(Surv(start, survt, status) ~ prison + dose + Clinic + logt_dose + cluster(ID), data = _, method = "efron") |> 
  summary()
```
```{r}
addicts |> 
  mutate(logt_dose = dose * log(survt)) |> 
  coxph(Surv(survt, status) ~ prison + dose + Clinic + logt_dose, data = _, method = "breslow") |> 
  summary()
```
### heaviside function

```{r}
addicts.cp365 <- addicts |>
  survSplit(
    Surv(survt, status) ~ .,
    data = _,
    cut = 365,
    event = "status",
    start = "start",
    end = "survt"
  ) |>
  mutate(hv1 = Clinic * (start < 365),
         hv2 = Clinic * (start >= 365))
```
```{r}
addicts.cp365 |> 
  coxph(Surv(start, survt, status) ~ prison + dose + hv1 + hv2 + cluster(ID), data = _, method = "breslow") |> 
  summary()
```
```{r}
vignette("survival")
vignette("timedep")
lung |> 
  mutate(time/30.5,
         .keep = "used") |> 
  pmin(12) |> 
  floor() 
```

```{r}
cgd
```
```{r}
cgd |> 
  coxph(Surv(tstart, tstop, status) ~ treat + inherit + steroids, cluster = id, data = _) |> 
  summary()
```

```{r}
cgd0
```

```{r}
jasa1
```

```{r}
jasa1 |> 
  coxph(Surv(start, stop, event) ~ age * transplant + surgery + year, data = _, method = "breslow") |>
  summary()
```
```{r}
veteran
```

