---
title: "Political Leadership Survival in the Aftermath of Coups or Overstays: From Illegitimate Ascent to Unexpected Exit"
format:
  jasa-pdf:
    keep-tex: true  
    linestretch: 1.75
    papersize: A4
    indent: true
    number-sections: true
    keywords: [Political survival, Coups, Overstays]

date: last-modified
author:
  - name: Zhu Qi
    affiliations:
      - name: University of Essex
        department: Department of Government
  
abstract: |
 Previous research predominantly focused on the disruption of regular leadership tenures by unexpected events, such as coups, self-coups, uprisings, rebellions, civil wars, or revolutions. In contrast, this study aims to delve into the longevity of leaders who ascend to power through these very unexpected events, specifically coups or overstays. The central argument posits that the endurance of political leadership is influenced not only by their actions and policies in office but also by the means through which they come to power. Employing a survival model, this research investigates the disparities in survival rates between leaders who rise to power via coups and those who overstay their terms, and seeks to elucidate the underlying reasons for these differences.

bibliography: survival.bib
mainfont: Times New Roman
fontsize: 12pt
execute: 
  freeze: auto
---

## Introduction

The question of why some leaders maintain their hold on power for extended periods, spanning decades, while others witness their leadership cut short after mere years, months, or even days, has captivated scholars and researchers in the field of political science.
This inquiry has been extensively explored in numerous works, as evidenced by notable contributions such as those by @clinton1975politics and @buenodemesquita2003.

In their seminal work, Bueno de Mesquita and his colleagues undertake a comprehensive examination of leaders across a diverse political landscape, encompassing democracies and autocracies, parliamentary and presidential systems, and both civilian and military contexts.
However, it is worth noting that a significant number of political leaders, especially in democratic countries, adhere to regular and predictable tenures.
An illustrative example can be found in the United States, where presidents may serve up to eight years if they secure a second term, even in cases of suboptimal performance.
Similarly, in autocratic Mexico from 1919 to 2000, each president served a fixed six-year term without facing overthrows or overstays.
In such contexts, the investigation of tenure length is of marginal significance, as power transitions between leaders typically occur within the established framework of constitutional rules or unwritten conventions.

Given the distinctive nature of political leader survival in different regimes, scholars have increasingly focused on the unexpected tenures, namely those leaders who do not complete their original terms or those who overstay their mandates.
This shift in focus stems from the fact that some leaders are toppled by coups, uprisings, rebellions, civil wars, or revolutions, while others successfully navigate lawful or unlawful challenges.
Previous research on the longevity of political leaders predominantly centers on two primary dimensions.
The first dimension encompasses the contextual conditions and resources available to leaders, including factors such as their personal competence [@yu2016], the stability of their society [@arriola2009], economic performance [@palmer1999, @williams2011], access to natural resources [@smith2004; @quirozflores2012], and external support networks [@licht2009; @wright2008, @thyne2017].
The second dimension delves into the strategies employed by leaders in enacting their political and economic policies [@gandhi2007; @morrison2009], as well as their responses to challenges and dissent within their regimes [@escribà-folch2013; @davenport2021].

Unsurprisingly, a substantial portion of the existing research on political survival predominantly centers on coups, as they represent the most common pathways to the exit of authoritarian leaders [@svolik2008; @frantz2016].
Previous literature has primarily delved into the survival of leadership in terms of strategies aimed at preventing coups [@powell2017; @sudduth2017; @debruin2020], or how leaders can extend their tenures after surviving failed coup attempts [@easton2018].

However, on one hand, the duration of political leaders' tenures can be significantly influenced by unforeseen events like coups.
On the other hand, these very unexpected events that usher in new leaders can also become the catalyst for the subsequent cycle of unexpected developments.
It is conceivable that leaders who come to power through regular and constitutional transitions are more likely to undergo periodic shifts in leadership, while those who seize power through unconstitutional means face a higher risk of unanticipated removal from office.
Unfortunately, there has been a limited emphasis on the study of leadership survival in the context of successful coups.
A similar research gap exists in the examination of incumbents who overstay their terms in power, which forms the central focus of this paper.

The analysis of their tenures holds particular significance for two reasons.
Firstly, the duration of these leaders' tenures exhibits considerable variation, ranging from mere months to several decades.
Secondly, predicting the tenures of such leaders proves challenging.
A seemingly robust and stable regime can collapse suddenly overnight, while an apparently fragile one might persist for decades.
The substantial disparities in these tenures remain inadequately explained, posing a perplexing challenge that has piqued the interest of numerous political scientists.

Expanding on the discourse surrounding coups and leaders who overstay their intended terms, this paper delves into the trajectories of political leaders who came to power through coups or extended their mandates beyond the originally intended tenure.
The central objective is to examine the variations in survival duration between leaders who attain power through coups and those who exceed their terms, while also shedding light on the underlying factors contributing to these distinctions.

This paper follows a structured approach as outlined below: The second section encompasses a comprehensive literature review on political survival and highlighting the contributions of this paper might offer.
The third chapter delves into the examination of factors influencing the survival of leaders who have ascended to power through unconstitutional means.
Chapter 4 provides an account of the methodology and data employed, utilizing a survival model for a comprehensive analysis of the determinants of leaders' survival.
The subsequent chapter, Chapter 5, presents the findings of this analysis, facilitating an in-depth discussion of the results.
Finally, in Chapter 6, the paper concludes by synthesizing these findings and exploring their broader implications.

## Literature review: The dynamics of leadership survival in different scenarios

In their ambitious work, @buenodemesquita2003 set out to provide a comprehensive explanation for the logic of political leadership survival within a universal framework.
According to this framework, the endurance of political leaders hinges on the maintenance of a supportive winning coalition.
However, it is essential to notice that the dynamics of leadership survival differ significantly across various types of regimes.

In democratic systems, distinctions emerge between parliamentary and presidential regimes.
For instance, in parliamentary countries such as the UK and Japan, political parties may maintain power for extended periods, even as prime ministers change frequently.
A contemporary example is the United Kingdom in 2022, which saw three different prime ministers while the Conservative Party retained its grip on power.
In contrast, in presidential countries like the United States, leaders are subject to fixed terms, making power transitions more regular and predictable.

Moreover, the concept of dividing the electorate into a selectorate and a winning coalition may not be as relevant in democracies.
While those who support and vote for incumbents may witness their preferred policies enacted, those who vote against them still experience the same policies.
For instance, individuals who cast their votes for candidates advocating lower taxes face the same tax rates as those who vote against the incumbents.
This doesn't result in lower taxes for supporters and higher taxes for opponents; rather, both groups encounter identical tax levels.
Consequently, it becomes challenging to argue that winning coalitions inherently gain a significant advantage over the broader selectorate in democratic systems.

On the other hand, there are types of autocratic regimes, each with its distinct characteristics, including civilian autocracy, personnel autocracy, military regimes, party dominance, and monarchies.
In most autocratic regimes, the process of leadership selection remains veiled in secrecy.
For instance, in countries like China, the mechanisms for appointing leaders often resemble a black box, concealing the rules and procedures from outsiders.
Expressing dissenting views, whether as potential challengers or supporters of challengers, can be perilous.
In Russia, despite the presence of general elections, challengers frequently face severe consequences such as assassination, poisoning, imprisonment, or exile.
As a result of the absence of transparent and fair conventional procedures for power transitions, leaders in autocratic regimes are more vulnerable to being deposed through coups or other unconventional means.

Beyond distinctions among various regimes, the endurance of leaders can fluctuate even within the confines of a single regime, contingent on the circumstances they encounter.
It stands to reason that leaders ascending through conventional means may experience a different survival trajectory compared to those ushered in by coups or those who overstay their terms.
Additionally, leaders operating in favorable economic, social, and international contexts are likely to have a more prolonged tenure compared to their counterparts navigating challenging conditions.

Considering the factors discussed earlier, a substantial portion of existing literature seeks to unravel the underlying principles governing political survival in non-democratic regimes.
Notably, scholarly attention has gravitated towards the examination of coup-proofing strategies, given that coups emerge as a primary driver of irregular exits in autocracies [@quinlivan1999; @powell2014; @sudduth2017; @tang2021].
Additionally, there is a notable focus on the study of survival strategies following failed coup attempts, as evidenced by the works of [@kebschull1994; @easton2018; @oztig2020].

In @sudduth2017a, the author delves into the post-coup actions of a dictator, despite the primary focus of the paper being on purge strategies.
The central argument asserts that leaders who rise to power through coups experience a temporary surge in influence compared to the elites immediately following the coups, making them less susceptible to being ousted by subsequent coup attempts.
This assertion, as highlighted by the author, challenges the conventional notion that new leaders are generally in a position of weakness in the initial stages of their tenure [@roessler2011].
Regardless of their initial strength, both Sudduth and previous scholars concur that new leaders are inclined to purge rival elite groups to bolster their power.
The distinction lies in Sudduth's claim that dictators undertake purges when they possess the capability to do so without significant risk, while conventional views posit that dictators resort to purges when compelled to prevent potential ousting by coups.

Yet, it's important to recognize that new leaders, especially those who ascend through unconventional means, don't conform to a universal pattern of being either inherently weak or powerful.
Leadership transitions occur in diverse contexts, and thus, leaders face a spectrum of challenges.
Some emerge in positions of vulnerability, while others wield considerable strength.
Regardless of individual power, when juxtaposed with the entirety of elites or the entire population, leaders remain in a position of relative weakness---unity among elites or residents can overshadow even the most powerful leaders.

## The logic of political leader survival in irregular ascensions

Engaging in a discourse on the survival strategies of political leaders within non-democratic regimes presents a significant challenge.
The complexity stems from the lack of a universal pattern that encapsulates the rules or conventions dictating power transitions in autocratic systems.
For instance, even in Middle Eastern monarchies, the transfer of power doesn't rigidly adhere to a father-to-son lineage.
However, this doesn't imply that analyzing survival strategies in autocracies is unattainable.
Despite substantial differences, they share certain commonalities.
Most autocratic regimes, especially those characterized by irregular ascensions, exhibit three prevalent situations.

The first aspect concerns the issue of legitimacy.
Leaders who ascend through coups lack legitimacy as they seize power through force or other unconventional means.
While many leaders prolong their tenures through a façade of constitutional procedures, such as judgments by the Supreme Court, congressional votes, or even referendums, they often manipulate these processes to maintain control.
It's commonly understood by ruling elites, opposition parties, and the populace that these leaders lack legitimacy.
This lack of legitimacy can sometimes justify the cause of those seeking their replacement, even if the means used are unconstitutional.

The second characteristic revolves around the uncertainty surrounding power transitions.
This uncertainty creates ambiguity not only for ruling elites and ordinary citizens but also for the leaders themselves regarding when, how, and to whom power might be transferred.
Such uncertainty breeds inherent instability.
Amidst such instability, people experience a lack of security.
This perception often leads to the belief that the current ruler is incompetent and should be replaced by someone more powerful or capable.
Consequently, the ruling elite or opposition factions may exploit the instability as an opportunity to challenge existing power structures.

The third aspect revolves around the equilibrium of power.
In autocracies, governance typically rests with a minority faction that possesses a highly structured organization, distinctly contrasting with the decentralized and disorganized subjects.
Even amid protests or uprisings, ruling groups adeptly suppress such incidents individually.
The possibility of overthrowing tyranny arises if the subjects can coalesce their efforts.
However, the principal obstacle lies in the formidable challenge of surmounting the collective action problem.
Autocratic dictators commonly adopt a ruling strategy focused on preventing unity among subjects and complicating endeavors to address collective action issues.
This elucidates why dictatorships curtail free expression, assembly, and association.
The absence of free public expression and association renders the power balance unclear.
Consequently, rulers maintain relative power advantages over not only the elites but also the subjects.

The trifecta of illegitimacy, uncertainty, and the equilibrium of power profoundly impact the longevity of a regime.
Yet, compared to leaders who gain power through coups, those who overstay their terms find themselves in a comparatively advantageous position concerning these three aspects.

### Legitimacy

As per Powell and Thyne's definition, coups constitute "illegal and overt attempts by the military or other elites within the state apparatus to unseat the sitting executive." [@powell2011, p.252].
While it is undeniable that a few coups have been justified by resolving crises and leading to improved outcomes, they remain illegal means to remove incumbents.
These unlawful methods open Pandora's box, publicly suggesting alternatives to constitutional procedures for seizing power, particularly in the case of successful coups.
Such actions inevitably prompt imitators to launch new coups.
As society becomes accustomed to coups, subsequent ones may not elicit significant backlash, given that the incumbents themselves ascended to power through similar means.
Furthermore, coups not only invite further coups but also embolden external challengers, including uprisings, revolutions, and civil wars [@dahl2023].

On the other hand, leaders who overstay their tenures may lack legitimacy but often manage to maintain power through a facade of legitimacy.
They don't blatantly seize power via military force but rather cling to power through parliamentary or congressional processes, the Supreme Court, and even nationwide referendums.
The opposition usually chooses to confront these leaders using legal means, engaging in legislative debates or legal proceedings, and sometimes by advocating for another referendum.
Attempts to overthrow leaders who overstay their terms through coups would be even less legitimate and might struggle to garner support.
However, removing such leaders within the boundaries of the law presents an arduous challenge, if not a near-impossible task.

### Uncertainty

As previously discussed, both leaders who assume power through coups and those who extend their terms contribute to uncertainties within their regimes.
However, regimes led by overstaying rulers generally demonstrate lesser overall uncertainty.

Primarily, following coups, regimes face uncertainty regarding four potential outcomes: democracies initially overthrown by coups may either persist as democracies (I) or transition into autocracies (II), while autocracies overthrown by coups may either endure as autocracies (III) or evolve into democracies (IV).
In contrast, regimes with leaders overstaying their terms, with only rare exceptions, typically persist as autocracies.

Secondly, the anticipated duration of future ruling tenures tends to be longer for overstaying rulers than for coup-entry leaders.
Most overstaying leaders endeavor to prolong their tenures for as long as possible.
For example, figures like Putin in Russia and Xi Jinping in China are less inclined to voluntarily step down unless forcibly removed from power.
Conversely, coup-entry leaders face greater uncertainty.
Coups are often justified through diverse excuses or claims.
Some assert their actions as defending democratic order, exemplified by President Manuel Zelaya's ousting in Honduras in 2009.
Others claim to protect the constitution, such as President Mamadou Tandja's overthrow in Niger in 2010.
The complexity arises when certain coup leaders honor their promises by transferring power to a constitutional successor.
For instance, following the 2010 coup in Niger, a new constitution restored civilian power and reinstated a strict two-term limit on the presidency in the same year [@ginsburg2019].
However, numerous others refuse to step down and retain power, as observed in the 1973 coup in Chile [@ökten2022].

Thirdly, uncertainties abound regarding policy changes and power reshuffling.
Comparatively, coup-entry leaders need to restructure top officials and justify their actions by implementing noticeable differences after overthrowing the incumbents.
In contrast, overstaying leaders face fewer issues as their regimes experience minimal dramatic changes.
There is no rush to dismantle the old ruling paradigm and establish a new order.

### Equilibrium of power

Rulers in autocracies face a challenging dilemma in managing powerful elites.
They require a strong force to counter potential external threats while constantly grappling with the fear of being replaced by competent and ambitious subordinates.
Therefore, maintaining a delicate balance of power becomes a sophisticated skill.
It's apparent that sustaining equilibrium is easier than restoring it once disrupted.

In regimes with leaders overstaying their terms, there exists, at least temporarily, a superficial equilibrium of power.
Successful overstays demonstrate the incumbents' firm grip on power, effectively deterring both internal dissent and external challenges.
Consequently, this contributes to the overall stability of the governing structure and society.

Conversely, leaders who come to power through coups invariably disrupt the balance of power and must establish a new equilibrium, even in relatively peaceful coup scenarios.
The overthrow of previous rulers necessitates the dismantling of the existing ruling framework and a reshuffling of top officers and officials.
These actions inevitably breed turbulence and adversaries for the new rulers, making the restoration of order and the balance of power considerably more challenging.

Internal conflicts among elites, as highlighted by numerous scholars, pose a significant threat to the stability of those in power.
Hence, they are often compelled to negotiate and compromise among external factions.
As noted by @roessler2011, these rulers might attempt to reduce the likelihood of subsequent coups, albeit at the expense of increasing the risk of societal rebellions and civil wars.
This strategy resonates with the approach adopted by Chinese leader Chiang Kai-shek in the 1930s.
He pursued a strategy of compromise with both Japan and the Soviet Union to eliminate the internal threat posed by the Chinese Communist Party, prioritizing "Domestic stability takes precedence over resisting foreign invasion" during that period [@chu1999chiang].
However, adopting a hardline stance internally could not only weaken themselves by purging elites, but also incite backlash from close allies, as witnessed in instances like Idi Amin's coup in 1971 in Uganda and Pervez Musharraf's coup in 1999 in Pakistan [@sudduth2017a].

These factors collectively contribute to a shorter expected lifespan of regimes following coups [@dahl2023], compared to relatively longer tenures for overstaying regimes.
Analyzing a comprehensive coup dataset [@powell2011] spanning from 1950 to 2023, it's evident that 97 countries experienced coups during this period.
Among them, 15 countries endured at least 10 coups, and 10 countries witnessed more than 6 successful coups, reinforcing the analysis discussed above.
Based on these insights, I propose the following hypothesis:

```{r eval=FALSE, include=FALSE}
library(tidyverse)
coups <- read.csv("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt",sep = "\t")

attempts <- coups %>% 
  select(country) %>% 
  summarise(count=n(),.by = country) %>% 
  arrange(-count) 

success <- coups %>% 
  filter(coup==2) %>% 
  select(country) %>% 
  summarise(count=n(),.by = country) %>% 
  arrange(-count) 

sum(Sum$count[1:10])
```

```{r eval=FALSE, include=FALSE}
table(coups$coup)
```

**Hypothesis:** Political leaders who successfully extend their time in power are more likely to have prolonged survival compared to leaders who assume power through coups.

## Method and data

\newpage

## References
