---
title: "Political Leadership Survival in the Aftermath of Coups or Overstays: From Illegitimate Ascent to Unexpected Exit"
format:
  jasa-pdf:
    keep-tex: true  
    linestretch: 1.75
    papersize: A4
    indent: true
    number-sections: true
    keywords: [Political survival, Coups, Overstays]

date: last-modified
author:
  - name: Zhu Qi
    affiliations:
      - name: University of Essex
        department: Department of Government
  
abstract: |
 This paper delves into the duration of political leaders' tenures, focusing on two specific categories: leaders who come into power through coups and those who exceed their designated term limits. It argues that the length of political leadership tenures is not solely determined by their governing strategies but also by the means through which they ascend to power. By employing a survival model, this study demonstrates that leaders entering through coups generally have shorter tenures compared to those who surpass their term limits. The primary objective of this research is to determine, on average, the extended survival duration of overstaying leaders in power in comparison to coup-entry leaders, aiming to uncover the underlying factors contributing to these divergent tenures.

bibliography: survival.bib
mainfont: Times New Roman
fontsize: 12pt
execute: 
  freeze: auto
---

## Introduction

The query of what drives certain leaders to endure in power for decades, in stark contrast to others whose leadership abruptly ends within short spans of time, has been a focal point of political science.
This compelling inquiry has garnered extensive attention and analysis across various scholarly works, as elaborated upon in Chapter 2.

Previous research on the survival of political leaders shows two noteworthy gaps.
Firstly, scholars have predominantly focused on generalized frameworks, which either encompass all types of regimes or include all forms of autocracies.
However, the survival dynamics significantly differ not just between democracies and autocracies but also among various types of autocratic rule.
Secondly, existing studies have primarily centered on analyzing the likelihood of events causing irregular leader exits, such as coups, occurring within a country-year and the factors influencing these events.
However, there remains a dearth of discussion concerning the average duration of survival, whether longer or shorter, for specific types of leaders compared to others.

This paper delves into a comparative analysis between two distinct categories of political leaders: those who ascend to power through coups and those who extend their designated term limits.
Existing research has consistently shown that the duration of political leaders' tenures can be profoundly influenced by unforeseen occurrences such as coups.
Paradoxically, these unexpected events that usher in new leaders can subsequently become determinants for further unforeseen developments.
It's plausible that leaders who assume power through regular and constitutional transitions are more prone to periodic shifts in leadership, while those seizing power through unconstitutional means face higher risks of unanticipated removal from office.

I propose that the means by which political leaders acquire power significantly impacts the duration of their leadership.
The examination of irregular-entry or overstaying leaders' tenures holds particular significance for two reasons.
Firstly, the tenures of irregular-entry leaders exhibit substantial variability compared to those of regular-entry leaders, ranging from brief periods to several decades.
Secondly, forecasting the tenures of irregular-entry leaders presents a challenge.
A seemingly robust and stable regime might unexpectedly collapse overnight, while an apparently fragile one might endure for decades.

Employing a survival model, I provide empirical evidence suggesting that leaders who surpass their term limits generally enjoy longer tenures compared to those entering through coups.
Coup-entry leaders, in contrast to overstaying leaders, grapple with more significant challenges related to legitimacy, uncertainty, instability, and power-sharing.
These factors potentially curtail the duration of survival for coup-entry leaders.

This study could offer contributions on two fronts.
Firstly, it underscores that the duration of survival and unforeseen exits isn't solely determined by leaders' actions post-assuming power but is inherently influenced by their method of ascension to power.
Secondly, it provides empirical measurements for comparing the tenure duration of different types of leaders.

After the introduction, the second section encompasses a comprehensive literature review on political survival and highlighting the contributions of this paper might offer.
The third chapter delves into the examination of factors influencing the survival of leaders who have ascended to power through unconstitutional means.
Chapter 4 provides an account of the methodology and data employed, utilizing a survival model for a comprehensive analysis of the determinants of leaders' survival.
Chapter 5 presents the findings of this analysis, facilitating an in-depth discussion of the results.
Finally, in Chapter 6, the paper concludes by synthesizing these findings and exploring their broader implications.

## Literature review: The dynamics of leadership survival in different scenarios

In their seminal work, @buenodemesquita2003 undertake a comprehensive examination of leaders across a diverse political landscape, encompassing democracies and autocracies, parliamentary and presidential systems, and both civilian and military contexts.
They try to provide a comprehensive explanation for the logic of political leadership survival within a universal framework.
However, it is essential to notice that the dynamics of leadership survival differ significantly across various types of regimes.

In democratic systems, distinctions emerge between parliamentary and presidential regimes.
For instance, in parliamentary countries such as the UK and Japan, political parties may maintain power for extended periods, even as prime ministers change frequently.
A contemporary example is the United Kingdom in 2022, which saw three different prime ministers while the Conservative Party retained its grip on power.
In contrast, in presidential countries like the United States, leaders are subject to fixed terms, making power transitions more regular and predictable.
Also, there are even more different types of autocratic regimes, each with its distinct characteristics, including civilian autocracy, personnel autocracy, military regimes, party dominance, and monarchies.
There are strictly periodically power transition cases.
An illustrative example can be found in autocratic Mexico from 1919 to 2000, each president served a fixed six-year term without facing overthrows or overstays \*\*\*.
There are many dictators who rule for life and even pass power to their sons as in Syria and North Korea, or brothers as in Cuba.
Of course, there are more rulers ousted unexpectedly.

Moreover, the concept of dividing the residents into different coalitions or groups may not be as relevant in democracies compared to autocracies.
In democracies, while those who support and vote for incumbents may witness their preferred policies enacted, those who vote against them still experience the same policies.
For instance, a candidate who advocates lower tax wins the election.
This doesn't result in lower taxes for supporters of this candidate and higher taxes for their opponents; rather, both groups encounter identical tax levels.
Things are different in autocracies where different groups, depending on their status, relationships with the rulers, do benefit differently.

Furthermore, the rules of power transitions in democracies and autocracies are very different.
In most autocratic regimes, the process of leadership selection remains veiled in secrecy.
For instance, in countries like China, the mechanisms for appointing leaders often resemble a black box, concealing the rules and procedures from outsiders.
Expressing dissenting views, whether as potential challengers or supporters of challengers, can be perilous.
In Russia, despite the presence of general elections, challengers frequently face severe consequences such as assassination, poisoning, imprisonment, or exile.
As a result of the absence of transparent and fair conventional procedures for power transitions, it is difficult to measure who are more powerful or who have more supporters.
It is possible that a dictator could survive a quite long tenure with a very weak supporting, because no one know the balance of power is not in the dictator's favour.
It is also possible a powerful ruler was ousted by a small number of guards, because the army or other supporters of the ruler hesitate to fight for him due to the lack of information that they are actually in predominant status in balance of power.
So it is impossible to accurately calculate the selectorates or coalitions, like @buenodemesquita2003 did in their research.
In democracies, the challengers could publicly run for leadership without any fear.
It is easy to find out their supporting rates.
They could try to gain more supporters through public speeches, TV or news paper advertisements, or assemblies.
If the supporting rates are not good enough to challenge the incumbents, different groups could try to cooperate.
So the power transitions could happen with a tiny more supports.

Given the distinctive nature of political leader survival in different regimes, more scholars have focused on logic of survival of specific types of regimes, delving into democracies (Svolik 2014 Which Democracies Will Last?) or autocracies (Davenport, RezaeeDaryakenari, and Wood 2021 Tenure Through Tyranny? Repression, Dissent, and Leader Removal in Africa and Latin America).
Of course, the unexpected tenures, namely those leaders who do not complete their original terms or those who overstay their terms, gain more academic attention.

Previous research on the longevity of political leaders predominantly centers on two primary dimensions.
The first dimension encompasses the contextual conditions and resources available to leaders (**objective conditions**), including factors such as their personal competence [@yu2016], the stability of the society [@arriola2009], economic development level [@palmer1999, @williams2011], natural resources [@smith2004; @quirozflores2012], and external support networks [@licht2009; @wright2008, @thyne2017].
The second dimension delves into the strategies employed by leaders in enacting their political and economic policies (**subjective actions**) [@gandhi2007; @morrison2009], as well as their responses to challengers and dissent within their regimes [@escribà-folch2013; @davenport2021].

Unsurprisingly, a substantial portion of these studies predominantly centers on coups, as they represent the most common pathways to the exit of authoritarian leaders [@svolik2008; @frantz2016].
Previous literature has primarily delved into the survival of leadership in terms of strategies aimed at preventing coups [@powell2017; @sudduth2017; @debruin2020], or how leaders can extend their tenures after surviving failed coup attempts [@easton2018].

Considering the factors discussed earlier, a substantial portion of existing literature seeks to unravel the underlying principles governing political survival in non-democratic regimes.
Notably, scholarly attention has gravitated towards the examination of coup-proofing strategies, given that coups emerge as a primary driver of irregular exits in autocracies [@quinlivan1999; @powell2014; @sudduth2017; @tang2021].
Additionally, there is a notable focus on the study of survival strategies following failed coup attempts [@kebschull1994; @easton2018; @oztig2020].

In @sudduth2017a, the author delves into the post-coup actions of a dictator, despite the primary focus of the paper being on purge strategies.
The central argument asserts that leaders who rise to power through coups experience a temporary surge in influence compared to the elites immediately following the coups, making them less susceptible to being ousted by subsequent coup attempts.
This assertion, as highlighted by the author, challenges the conventional notion that new leaders are generally in a position of weakness in the initial stages of their tenure [@roessler2011].
Regardless of their divergence, both Sudduth and previous scholars concur that new leaders are inclined to purge rival elite groups to bolster their power.
The distinction lies in Sudduth's claim that dictators undertake purges when they possess the capability to do so without significant risk, while conventional views posit that dictators resort to purges when compelled to act to immenent threats.

Yet, it's important to recognize that new leaders, especially those who ascend through unconventional means, don't conform to a universal pattern of being either inherently weak or powerful.
Leadership transitions occur in diverse contexts, and thus, leaders face different levels of challenges.

Academics have thoroughly examined the endurance of political leaders across diverse contexts, encompassing universal frameworks, autocratic regimes, and in the aftermath of failed challenges like coup attempts.
Surprisingly, there has been limited discussion regarding the survival dynamics of leaders who manage to retain power after successful coups or prolonged tenures of incumbents.
This paper seeks to address this gap by specifically examining and distinguishing the duration of survival between these two categories of leaders.

## The logic of political leader survival in irregular ascensions

Engaging in a discourse on the survival strategies of political leaders within non-democratic regimes presents a significant challenge.
The complexity stems from the lack of a universal pattern that encapsulates the rules or conventions dictating power transitions in autocratic systems.
For instance, even in Middle Eastern monarchies, the transfer of power doesn't rigidly adhere to a father-to-son lineage.
However, this doesn't imply that analyzing survival strategies in autocracies is unattainable.
Despite substantial differences, they share certain commonalities.
Most autocratic regimes, especially those characterized by irregular ascensions, exhibit three prevalent situations.

The first aspect concerns the issue of legitimacy.
Leaders who ascend through coups lack legitimacy as they seize power through force or other unconventional means.
While many leaders prolong their tenures through a façade of constitutional procedures, such as judgments by the Supreme Court, congressional votes, or even referendums, they often manipulate these processes to maintain control.
It's commonly understood by ruling elites, opposition parties, and the populace that these leaders lack legitimacy.
This lack of legitimacy can sometimes justify the cause of those seeking their replacement, even if the means used are unconstitutional.

The second characteristic revolves around the uncertainty surrounding power transitions.
This uncertainty creates ambiguity not only for ruling elites and ordinary citizens but also for the leaders themselves regarding when, how, and to whom power might be transferred.
Such uncertainty breeds inherent instability.
Amidst such instability, people experience a lack of security.
This perception often leads to the belief that the current ruler is incompetent and should be replaced by someone more powerful or capable.
Consequently, the ruling elite or opposition factions may exploit the instability as an opportunity to challenge existing power structures.

The third aspect revolves around the equilibrium of power.
In autocracies, governance typically rests with a minority faction that possesses a highly structured organization, distinctly contrasting with the decentralized and disorganized subjects.
Even amid protests or uprisings, ruling groups adeptly suppress such incidents individually.
The possibility of overthrowing tyranny arises if the subjects can coalesce their efforts.
However, the principal obstacle lies in the formidable challenge of surmounting the collective action problem.
Autocratic dictators commonly adopt a ruling strategy focused on preventing unity among subjects and complicating endeavors to address collective action issues.
This elucidates why dictatorships curtail free expression, assembly, and association.
The absence of free public expression and association renders the power balance unclear.
Consequently, rulers maintain relative power advantages over not only the elites but also the subjects.

The trifecta of illegitimacy, uncertainty, and the equilibrium of power profoundly impact the longevity of a regime.
Yet, compared to leaders who gain power through coups, those who overstay their terms find themselves in a comparatively advantageous position concerning these three aspects.

### Legitimacy

As per Powell and Thyne's definition, coups constitute "illegal and overt attempts by the military or other elites within the state apparatus to unseat the sitting executive." [@powell2011, p.252].
While it is undeniable that a few coups have been justified by resolving crises and leading to improved outcomes, they remain illegal means to remove incumbents.
These unlawful methods open Pandora's box, publicly suggesting alternatives to constitutional procedures for seizing power, particularly in the case of successful coups.
Such actions inevitably prompt imitators to launch new coups.
As society becomes accustomed to coups, subsequent ones may not elicit significant backlash, given that the incumbents themselves ascended to power through similar means.
Furthermore, coups not only invite further coups but also embolden external challengers, including uprisings, revolutions, and civil wars [@dahl2023].

On the other hand, leaders who overstay their tenures may lack legitimacy but often manage to maintain power through a facade of legitimacy.
They don't blatantly seize power via military force but rather cling to power through parliamentary or congressional processes, the Supreme Court, and even nationwide referendums.
The opposition usually chooses to confront these leaders using legal means, engaging in legislative debates or legal proceedings, and sometimes by advocating for another referendum.
Attempts to overthrow leaders who overstay their terms through coups would be even less legitimate and might struggle to garner support.
However, removing such leaders within the boundaries of the law presents an arduous challenge, if not a near-impossible task.

### Uncertainty

As previously discussed, both leaders who assume power through coups and those who extend their terms contribute to uncertainties within their regimes.
However, regimes led by overstaying rulers generally demonstrate lesser overall uncertainty.

Primarily, following coups, regimes face uncertainty regarding four potential outcomes: democracies initially overthrown by coups may either persist as democracies (I) or transition into autocracies (II), while autocracies overthrown by coups may either endure as autocracies (III) or evolve into democracies (IV).
In contrast, regimes with leaders overstaying their terms, with only rare exceptions, typically persist as autocracies.

Secondly, the anticipated duration of future ruling tenures tends to be longer for overstaying rulers than for coup-entry leaders.
Most overstaying leaders endeavor to prolong their tenures for as long as possible.
For example, figures like Putin in Russia and Xi Jinping in China are less inclined to voluntarily step down unless forcibly removed from power.
Conversely, coup-entry leaders face greater uncertainty.
Coups are often justified through diverse excuses or claims.
Some assert their actions as defending democratic order, exemplified by President Manuel Zelaya's ousting in Honduras in 2009.
Others claim to protect the constitution, such as President Mamadou Tandja's overthrow in Niger in 2010.
The complexity arises when certain coup leaders honor their promises by transferring power to a constitutional successor.
For instance, following the 2010 coup in Niger, a new constitution restored civilian power and reinstated a strict two-term limit on the presidency in the same year [@ginsburg2019].
However, numerous others refuse to step down and retain power, as observed in the 1973 coup in Chile [@ökten2022].

Thirdly, uncertainties abound regarding policy changes and power reshuffling.
Comparatively, coup-entry leaders need to restructure top officials and justify their actions by implementing noticeable differences after overthrowing the incumbents.
In contrast, overstaying leaders face fewer issues as their regimes experience minimal dramatic changes.
There is no rush to dismantle the old ruling paradigm and establish a new order.

### Equilibrium of power

Rulers in autocracies face a challenging dilemma in managing powerful elites.
They require a strong force to counter potential external threats while constantly grappling with the fear of being replaced by competent and ambitious subordinates.
Therefore, maintaining a delicate balance of power becomes a sophisticated skill.
It's apparent that sustaining equilibrium is easier than restoring it once disrupted.

In regimes with leaders overstaying their terms, there exists, at least temporarily, a superficial equilibrium of power.
Successful overstays demonstrate the incumbents' firm grip on power, effectively deterring both internal dissent and external challenges.
Consequently, this contributes to the overall stability of the governing structure and society.

Conversely, leaders who come to power through coups invariably disrupt the balance of power and must establish a new equilibrium, even in relatively peaceful coup scenarios.
The overthrow of previous rulers necessitates the dismantling of the existing ruling framework and a reshuffling of top officers and officials.
These actions inevitably breed turbulence and adversaries for the new rulers, making the restoration of order and the balance of power considerably more challenging.

Internal conflicts among elites, as highlighted by numerous scholars, pose a significant threat to the stability of those in power.
Hence, they are often compelled to negotiate and compromise among external factions.
As noted by @roessler2011, these rulers might attempt to reduce the likelihood of subsequent coups, albeit at the expense of increasing the risk of societal rebellions and civil wars.
This strategy resonates with the approach adopted by Chinese leader Chiang Kai-shek in the 1930s.
He pursued a strategy of compromise with both Japan and the Soviet Union to eliminate the internal threat posed by the Chinese Communist Party, prioritizing "Domestic stability takes precedence over resisting foreign invasion" during that period [@chu1999chiang].
However, adopting a hardline stance internally could not only weaken themselves by purging elites, but also incite backlash from close allies, as witnessed in instances like Idi Amin's coup in 1971 in Uganda and Pervez Musharraf's coup in 1999 in Pakistan [@sudduth2017a].

These factors collectively contribute to a shorter expected lifespan of regimes following coups [@dahl2023], compared to relatively longer tenures for overstaying regimes.
Analyzing a comprehensive coup dataset [@powell2011] spanning from 1950 to 2023, it's evident that 97 countries experienced coups during this period.
Among them, 15 countries endured at least 10 coups, and 10 countries witnessed more than 6 successful coups, reinforcing the analysis discussed above.
Based on these insights, I propose the following hypothesis:

```{r eval=FALSE, include=FALSE}
library(tidyverse)
coups <- read.csv("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt",sep = "\t")

attempts <- coups %>% 
  select(country) %>% 
  summarise(count=n(),.by = country) %>% 
  arrange(-count) 

success <- coups %>% 
  filter(coup==2) %>% 
  select(country) %>% 
  summarise(count=n(),.by = country) %>% 
  arrange(-count) 

sum(Sum$count[1:10])
```

```{r eval=FALSE, include=FALSE}
table(coups$coup)
```

**Hypothesis:** Political leaders who successfully extend their time in power are more likely to have prolonged survival compared to leaders who assume power through coups.

## Method and data

\newpage

## References
